---
title: "PA1_Template"
author: "Unnat Palan"
date: "April 15, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading activity.csv file
```{r reading the data}
activity <- read.csv("activity.csv", sep = ",")
```
Removing rows with NA
```{r removing NA}
names(activity)
str(activity)
head(activity[which(!is.na(activity$steps)), ])
```
Plotting the Histogram of the total number of steps taken each day
```{r histogram for daily steps}
library(reshape2)
activity_m <- melt(activity[which(!is.na(activity$steps)), ], id.vars = c("date", "interval"))
steps_sum <- dcast(activity_m, date ~ variable, sum)
hist(steps_sum$steps, main = "Histogram of total steps taken per day", 
     xlab = "Total steps per day", ylab = "Number of days", 
     breaks = 10, col = "green")
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"), 
       lty = c(1, 2), lwd = c(2, 2))
```

Calculating the mean of the total number of steps taken per day
```{r steps mean}
library(reshape2)
activity_m <- melt(activity[which(!is.na(activity$steps)), ], id.vars = c("date", "interval"))
head(activity_m)
steps_sum <- dcast(activity_m, date ~ variable, sum)
head(steps_sum)
```

Time series plot of the average number of steps taken
```{r average steps taken}
stepsdailyaverage <- dcast(activity_m, interval ~ variable, 
                           mean, na.rm = TRUE)
head(stepsdailyaverage)
plot(stepsdailyaverage$interval, stepsdailyaverage$steps, ty = "l",
     xlab = "time interval", ylab = "Average steps", 
     main = "Average steps taken over all days vs \n time interval")
```

5-minute interval that, on average, contains the maximum number of steps
```{r max steps}
maxsteps_interval <- 
  stepsdailyaverage$interval[which.max(stepsdailyaverage$steps)]
maxsteps_interval
```

Imputing missing data
```{r imputing missing data}
activity2 <- split(activity, activity$interval)

activity2 <- lapply(activity2, function(x) {
  x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
  return(x)
})

activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL

activity2 <- split(activity2, activity2$date)

df <- lapply(activity2, function(x) {
  x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
  return(x)
})

activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL
head(activity2)

library(reshape2)
activity_m2 <- melt(activity2, id.vars = c("date", "interval"))
steps_sum <- dcast(activity_m2, date ~ variable, sum, na.rm = TRUE)
head(steps_sum)
```

Plotting the Histogram of the total number of steps taken each day after missing values are imputed
```{r histogram with imputed data}
hist(steps_sum$steps, main = "Histogram of total steps taken per day", 
     xlab = "Total steps per day", ylab = "Number of days", 
     breaks = 10, col = "grey")
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"), 
       lty = c(2, 1), lwd = c(2, 2))
```

Comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r average steps across weekdays and weekends} 
weekends <- which(weekdays(as.Date(activity2$date)) == "Saturday" |
                    weekdays(as.Date(activity2$date)) == "Sunday")
weekdays <- which(weekdays(as.Date(activity2$date)) != "Saturday" &
                    weekdays(as.Date(activity2$date)) != "Sunday")
temp <- c(rep("a", length(activity2)))
temp[weekends] <- "weekend"
temp[weekdays] <- "weekday"
length(temp)
activity2 <- cbind(activity2, temp)
head(activity2)
names(activity2)[4] <- "day"

activity2split <- split(activity2, activity2$day)

stepsmean_interval <- lapply(activity2split, function(x) {
  temp <- aggregate(x$steps, list(x$interval), mean)
  names(temp) <- c("interval", "steps")
  return(temp)
})

stepsmean_interval <- do.call("rbind", stepsmean_interval)
weekdays <- grep("weekday" ,row.names(stepsmean_interval))
weekends <- grep("weekend" ,row.names(stepsmean_interval))
temp <- c(rep("a", length(stepsmean_interval$steps)))
temp[weekdays] <- "weekdays"
temp[weekends] <- "weekends"
names(temp) <- "day"
stepsmean_interval <- cbind(stepsmean_interval, temp)
row.names(stepsmean_interval) <- NULL

head(stepsmean_interval)

library(ggplot2)
ggplot(stepsmean_interval, aes(interval, steps)) + geom_line() + facet_grid(temp ~ .) 
```
